---
title: Python code!! 
author: Fiona MacNeill
date: '2019-12-13'
slug: bloggy-blog
categories: []
tags: []
description: ''
---
##Here are some neat things that we've been doing in python!! 

```{r}
###First I'm gonna load reticulate 
library(reticulate)

```

##Here's some basic function coding in python- let's say you start with a list of numbers or letters and way to do a simple loop printing them, you could do something like this! 

```{python}
for letter in "pythoniscool":
    print(letter)
```

```{python}
primes=[2,3,5,7,11]
for i in primes:
    print(i)
```
##You can also use "for" commands like this: 
```{python}
for i in range (1,30,3):
    print(i)###I just told python to start at 1 and count up to 30 by intervals of 3
```


##You can count occurances in a string like this: 
```{python}

import numpy as np

import pandas as pd
```

```{python}
def count_chars(sentence): 
    counts = {} #make an empty dictionary
    for c in sentence: #for every value in the sentence 
        if c in counts:
            counts[c]+=1 #if its a value that's already shown up add +1 to counts
        else:
            counts[c]=1 #If it's a new value, hold counts for values steady
    for c in counts:
        print(c, "appears", counts[c],"times")##Telling what to put in between characters, for example 'c appears 2 times' etc. 


##Let's try it out!! 
sentence1= "I can learn to code" 
count_chars(sentence1)
```
##Ok so we can see how often a single character appears in a group, but we can also see how many times GROUPS of characters appear with a (similar but modified) function. Let's look at nucleotide sequences as an example 

```{python}
def kmers(read, k):  ##defining a function for some sequecnce you're reading 'read' by groups 'k'. For our example k=3 
    counts = {}  ##Another empty dictionary 
    num_kmers = len(read) - k + 1###sets the numbers to read length of the seuqnce -one more than the selected length of our read 
    for i in range(num_kmers):
        kmer = read[i:i+k]##reading from 1 through however many we picked (so for example, 1-4 gets you a read of 3)
        if kmer not in counts:##This count function is pretty much exactly the same!
            counts[kmer] = 0
        counts[kmer] += 1
    return counts
##Let's try it!

my_seq1= "ATCATGATCATGATCATC" 

kmers(my_seq1,3)

##If we wanted to count by 4, we could just do this 

kmers(my_seq1,4) 
```

##Some basic regex 

```{python}
import re

string="I can learn to code" 
match=re.match(r"I.*learn",string) ##Gives us everything in the string between I&Learn 
match.group()


##Maybe we just want the numbers out of a string 

String2= "In a string like this maybe I just want 1, 2, 3, and 4, but I also want 777, and 58" 

##I'll use re.findall to pull out the numbers 

re.findall(r'\d+', String2)

```


```{python}
string2= "Jan 13 00:48:59: DROP service 68->67(udp) from 213.92.153.167 to 69.43.107.219, prefix: \"spoof iana-0/8\" \
(in: eth0 69.43.112.233(38:f8:b7:90:45:92):68 -> 217.70.100.113(00:21:87:79:9c:d9):67 UDP len:576 ttl:64) \
Jan 13 12:02:48: ACCEPT service dns from 74.125.186.208 to firewall(pub-nic-dns), prefix: \"none\" \
(in: eth0 74.125.186.208(00:1a:e3:52:5d:8e):36008 -> 140.105.63.158(00:1a:9a:86:2e:62):53 UDP len:82 ttl:38) \
Jan 13 17:44:52: DROP service 68->67(udp) from 172.45.240.237 to 217.70.177.60, prefix: \"spoof iana-0/8\" \
(in: eth0 216.34.90.16(00:21:91:fe:a2:6f):68 -> 69.43.85.253(00:07:e1:7c:53:db):67 UDP len:328 ttl:64) \
Jan 13 17:52:08: ACCEPT service http from 213.121.184.130 to firewall(pub-nic), prefix: \"none\" \
(in: eth0 213.121.184.130(00:05:2e:6a:a4:14):8504 -> 140.105.63.164(00:60:11:92:ed:1b):80 TCP flags: ****S* len:52 ttl:109)"


##Maybe we just want to pull out Mac IP assresses, which include numbers and special characters and letters 
##You can use re.finall all to recognize specific segments which indicate a Mac IP address 
re.findall(r'[0-9A-Fa-f]{2}\:[0-9A-Fa-f]{2}\:[0-9A-Fa-f]{2}\:[0-9A-Fa-f]{2}\:[0-9A-Fa-f]{2}\:[0-9A-Fa-f]{2}',string2)

```

##Using code in both python and R 


```{r}
library(reticulate)
hi<-"Hello"

thanks<-"Thank you"

```


```{python}

hi="World"

thanks="for reading!" 

print(r.hi,hi)
print(r.thanks,thanks)
```







